A comparison of single-cell trajectory inference methods: towards more accurate and robust tools

### 2018-03-01 analysis/paper/paper_latex.Rmd
update paper

```{r}
add_sfig(
figure_file("methods_timeline.pdf", "4-method_characterisation"),
"methods_over_time", 
"New TI methods over time. **a** Number of methods published or in preprint. **b** Number of methods which can handle a particular type of trajectory."
)
```

@@

# Results
## Method quality control

We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` TI methods  (`r ref("fig", "evaluation_overview", "a")`), and selected a subset of `r sum(methods$evaluated)` for evaluation, primarily based on their free availability, the presence of a programming interface, and the time they were published (`r ref("table", "implementations")`). We characterised all methods in four different ways: possible trajectory topologies, implementation quality, prior information and the underlying algorithm (`r ref("fig", "implementation_overview")`).

@@

`r anchor("fig", "evaluation_overview")`
![](`r figure_file("ti_evaluation_overview_v2.pdf", experiment_id="evaluation_overview")`)
`r ref("fig", "evaluation_overview")` Overview of the evaluation

@@

`r anchor("fig", "implementation_overview")`
![](`r figure_file("implementation_overview.pdf", experiment_id="4-method_characterisation")`)
`r ref("fig", "implementation_overview", anchor=TRUE)` First characterization of each method.

@@

```{r}
add_sfig(
figure_file("qc_over_time.pdf","4-method_characterisation"), 
"qc_over_time", 
"Quality control scores and number of citations over time"
)
```

@@

```{r}
add_sfig(
figure_file("method_components.pdf","4-method_characterisation"), 
"method_components", 
"TI methods frequently share steps"
)
```
## Evaluating the accuracy of the inferred trajectory

We compared the accuracy of each method by comparing its output with a gold standard (`r ref("fig", "evaluation_overview", "b")`). To make the heterogenous output of each method comparable, we created wrappers around each method so that it was converted into a common probabilistic model (`r ref("fig", "evaluation_overview", "e")` and `r ref("sfig", "wrapping")`. In this model the overall topology is represented by a network between "milestones", and the cells are placed within the space formed by each set of connected milestones (`r ref("fig", "evaluation_overview", "c")`). This common trajectory format allowed us to compare the output of each method against gold standard trajectories. We compared the results of each TI method with both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance. These synthetic datasets were generated using a realistic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and for each synthetic dataset a real dataset was used as reference to match the statistical properties of both datasets as best as possible (`r ref("fig", "evaluation_overview", "d")`). We used both large and smaller real datasets, from a variety of platforms, organisms and containing all different trajectory types (`r ref("fig", "evaluation_overview", "f")`). We further divided the datasets into gold and silver standard, depending on whether the gold standard was extracted from only external knowledge (gold) or including the expression data itself (silver). To compare the predictions of a method with a gold or silver standard, we developed three different score metrics, each assessing a different aspect of the trajectory: the cell ordering, the cell neighbourhood and topology (`r ref("fig", "evaluation_overview", "f")`). To calculate a final ordering, we combined metrics and datasets so that each metric, trajectory type and dataset source had equal weight.

```{r}
add_sfig({
        figure_file('wrapping.pdf', '4-method_characterisation')
}, 
"wrapping", 
"Our wrapping workflow", 
15, 
7.5
)
```

@@

```{r}
add_sfig(
        figure_file('real_synthetic_comparison.pdf', "5-optimise_parameters/5-real_synthetic_comparison")
, 
"real_synthetic_comparison", 
"Comparison of method performance between real and synthetic datasets. ", 
12, 
4
)
```
# Discussion
<!-- Some intro -->
Trajectory inference is unique among most other categories of single-cell analysis methods, such as clustering, normalisation and differential expression, because it models the data in a way that was almost impossible using bulk data. Indeed, for no other single-cell analysis types have so many tools been developed, according to omictools.org [@henry_omictoolsinformativedirectory_2014] (https://omictools.com/single-cell-rna-seq-category), the "awesome single cell software" list created by Sean Davis [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools [@zappia_exploringsinglecellrnaseq_2017] (https://www.scrna-tools.org) . It is therefore critical that these methods, now reaching `r sum(methods$is_ti, na.rm=TRUE)`, are evaluated to guide the users in their choice. In this preprint, we summarise the initial version of our evaluation of these methods, focusing on the quality control and the accuracy of their model using their default parameters on real and synthetic data.

@@

```{r}
add_sfig(
figure_file("trajectory_type_trees.pdf","2-dataset_characterisation/2-trajectory_types"), 
"trajectory_type_tree", 
"Different trajectory types can be classified in a directed acyclic graph, in which certain types are a generalisation of other trajectory types.", 
15, 
10
)
```

@@

```{r}
add_sfig({
figure_file("edge_flip_comparison.pdf", experiment_id="3-score_rules/4-edge_flip")
}, 
"edge_flip_comparison", 
"Edge flip score values for common trajectory topologies", 
15, 
5
)
```

@@

```{r}
git2r::repository(".")
```
I think we should mention something here about sc transcriptomics, flow/mass cytometry etc...
For example, I talked with Abhishek the other day and his view was really flow oriented. I think people who primarily use flow will be confused by whether this paper looks at scRNAseq or both flow and scRNAseq
Love this figure! It's a great summary of the /whole/ study methodology
I have several comments (that I might address myself if I get to it first). I'll post them as separate comments, so they can get resolved individually.
I would include a part for the method wrappers; that is the way that linear, cellcluster, cellgraph and generic TI's get translated to the probabilistic trajectory model
part c: I would explicitly show that the probabilistic trajectory models are being used for both the tasks and for the method outputs
The spacing is a bit variable; e.g. part b takes up a lot of real estate while parts d and e seem to be packed very tightly together
part f: I'd explicitly mention that each column is to be interpreted separately; there is no connection horizontally. JUST TO MAKE SURE.
a handful of the most recent methods
Repetition of most: put "The majority of" instead?
..

>>>>

```{r}
add_sfig(
figure_file("methods_timeline.tmp.pdf", "4-method_characterisation"),
"methods_over_time", 
"New TI methods over time. **a** Number of methods published or in preprint. **b** Number of methods which can handle a particular type of trajectory."
)
```

@@

# Results

We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` TI methods  (`r ref("fig", "evaluation_overview", "a")`), and selected a subset of `r sum(methods$evaluated)` for evaluation, primarily based on their free availability, the presence of a programming interface, and the time they were published (`r ref("table", "implementations")`). Currently, only methods published before june 2017 are included in the current evaluation, while more recent methods will be included in the next version of the evaluation. We characterised and benchmarked each method in six different ways: its detectable trajectory topology, its performance on real and synthetic datasets, its running time and scalability, its requirements for prior information, the underlying algorithm, and the quality of the tools. Based on these results, we develop a set of guidelines both for the method users, as well as for method developers.

## Performance on real and synthetic data
We compared the accuracy of each method by comparing its output with a gold standard (`r ref("fig", "evaluation_overview", "b")`). To make the heterogenous output of each method comparable, we created wrappers around each method so that it was converted into a common probabilistic model (`r ref("fig", "evaluation_overview", "e")` and `r ref("sfig", "wrapping")`. In this model the overall topology is represented by a network between "milestones", and the cells are placed within the space formed by each set of connected milestones (`r ref("fig", "evaluation_overview", "c")`). This common trajectory format allowed us to compare the output of each method against gold standard trajectories. We compared the results of each TI method with both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance. These synthetic datasets were generated using a realistic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and for each synthetic dataset a real dataset was used as reference to match the statistical properties of both datasets as best as possible (`r ref("fig", "evaluation_overview", "d")`). We used both large and smaller real datasets, from a variety of platforms, organisms and containing all different trajectory types (`r ref("fig", "evaluation_overview", "f")`). We further divided the datasets into gold and silver standard, depending on whether the gold standard was extracted from only external knowledge (gold) or including the expression data itself (silver). To compare the predictions of a method with a gold or silver standard, we developed three different score metrics, each assessing a different aspect of the trajectory: the cell ordering, the cell neighbourhood and topology (`r ref("fig", "evaluation_overview", "f")`). To calculate a final ordering, we combined metrics and datasets so that each metric, trajectory type and dataset source had equal weight.

```{r}
add_sfig({
        figure_file('wrapping.tmp.pdf', '4-method_characterisation')
}, 
"wrapping", 
"Our wrapping workflow", 
15, 
7.5
)
```

`r anchor("fig", "results_overview")`
![](`r figure_file("overview.tmp.pdf", experiment_id="9-main_figure")`)
`r ref("fig", "results_overview")` HERE ARE THE RESULTS

Overall, we find that slingshot predicts the most accurate trajectories across different data sources, trajectory types and metrics, followed by TSCAN and Monocle DDRTree CREF. These results are consistent between real and synthetic datasets CREF, confirming both the relevance of the synthetic data and the accuracy of the real datasets' gold and silver standards. However, results are more variable when comparing the three different evaluation metrics, 
## Detectable trajectory topologies

The main distinguishable feature among TI methods is the extent to which a method can adapt its trajectory topology to a dataset. In some 
## Prior information

## Running time and scalability

## Algorithm components

We characterised all methods in four different ways: possible trajectory topologies, implementation quality, prior information and the underlying algorithm (`r ref("fig", "implementation_overview")`).

@@

`r anchor("fig", "evaluation_overview")`
![](`r figure_file("ti_evaluation_overview_v2.tmp.pdf", experiment_id="evaluation_overview")`)
`r ref("fig", "evaluation_overview")` Overview of the evaluation

@@

`r anchor("fig", "implementation_overview")`
![](`r figure_file("implementation_overview.tmp.pdf", experiment_id="4-method_characterisation")`)
`r ref("fig", "implementation_overview", anchor=TRUE)` First characterization of each method.

@@

```{r}
add_sfig(
figure_file("qc_over_time.tmp.pdf","4-method_characterisation"), 
"qc_over_time", 
"Quality control scores and number of citations over time"
)
```

@@

```{r}
add_sfig(
figure_file("method_components.tmp.pdf","4-method_characterisation"), 
"method_components", 
"TI methods frequently share steps"
)
```
## Evaluating the accuracy of the inferred trajectory

@@

```{r}
add_sfig(
        figure_file('real_synthetic_comparison.tmp.pdf', "5-optimise_parameters/5-real_synthetic_comparison")
, 
"real_synthetic_comparison", 
"Comparison of method performance between real and synthetic datasets. ", 
12, 
4
)
```
# Discussion
<!-- Some intro -->
Trajectory inference is unique among most other categories of single-cell analysis methods, such as clustering, normalisation and differential expression, because it models the data in a way that was almost impossible using bulk data. Indeed, for no other single-cell analysis types have so many tools been developed, according to omictools.org [@henry_omictoolsinformativedirectory_2014] (https://omictools.com/single-cell-rna-seq-category), the "awesome single cell software" list created by Sean Davis [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools [@zappia_exploringsinglecellrnaseq_2017] (https://www.scrna-tools.org) . It is therefore critical that these methods, now reaching `r sum(methods$is_ti, na.rm=TRUE)`, are evaluated to guide the users in their choice. In this preprint, we summarise the initial version of our evaluation of these methods, focusing on the quality control and the accuracy of their model using their default parameters on real and synthetic data.

@@

```{r}
add_sfig(
figure_file("trajectory_type_trees.tmp.pdf","2-dataset_characterisation/2-trajectory_types"), 
"trajectory_type_tree", 
"Different trajectory types can be classified in a directed acyclic graph, in which certain types are a generalisation of other trajectory types.", 
15, 
10
)
```

@@

```{r}
add_sfig({
figure_file("edge_flip_comparison.tmp.pdf", experiment_id="3-score_rules/4-edge_flip")
}, 
"edge_flip_comparison", 
"Edge flip score values for common trajectory topologies", 
15, 
5
)
```

@@

```{r}
git2r::repository(".")
```
I think we should mention something here about sc transcriptomics, flow/mass cytometry etc...
For example, I talked with Abhishek the other day and his view was really flow oriented. I think people who primarily use flow will be confused by whether this paper looks at scRNAseq or both flow and scRNAseq
Love this figure! It's a great summary of the /whole/ study methodology
I have several comments (that I might address myself if I get to it first). I'll post them as separate comments, so they can get resolved individually.
I would include a part for the method wrappers; that is the way that linear, cellcluster, cellgraph and generic TI's get translated to the probabilistic trajectory model
This is complex though to put into this figure :p. Have a look at 4-method_characteristics/wrapping.tmp.pdf I think it is the least crucial of all parts of the figure now, so.. But feel free to try it yourself :p
a handful of the most recent methods
Repetition of most: put "The majority of" instead?
..
